A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2024-04-12 tidyindex.qmd
Fix typo intro sec 4
Changes sect 4

In what follows when geographic or temporal components of the $x_j(s,t)$ process are fixed we will be using suffix notation. For example, $x_{sj}(t)$ represents the data for a fixed location $s$ as a function of time $t$. While $x_{tj}(s)$ denotes the spatial varying process for a fixed $t$. An overview of the notation for pipeline input, operation, and output is present in @tbl-notation.

@@

The spatial processing module takes a single variable with a fixed temporal dimension, $x_{tj}(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]$. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and country scales.

@@

Variable transformation takes the input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T[.]$ to produce $x^{\text{Trans}}_{j}(s;t)$. When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality. For example, in the Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional shape of the variable. It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$. In the Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.

@@

Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^p$, or a subset of variables in $\mathbf{x}(s;t)$, as the input. It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $y \subseteq \mathbb{R}^d$ and $d < p$, as the output. The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@spei] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI), and can be considered to be a dimension reduction using a particular linear combination.

>>>>

In what follows when geographic or temporal components of the $x_j(s,t)$ process are fixed we will be using suffix notation. For example, $x_{sj}(t)$ represents the data for a fixed location $s$ as a function of time $t$. While $x_{tj}(s)$ denotes the spatial varying process for a fixed $t$. An overview of the notation for pipeline input, operation, and output is presented in @tbl-notation.

@@

The spatial processing module takes a single variable with a fixed temporal dimension, $x_{tj}(s)$, as input. This step transforms the variable from the original spatial dimension $s$ into the new dimension $s^\prime \in D_{s^\prime}$ through $x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]$ via a function $g$. The change of spatial dimension allows for the alignment of variables collected from different measurements, such as in-situ stations and satellite imagery, or originating from different resolutions. This also includes the aggregation of variables into different levels, such as city, state, and country scales.

@@

Variable transformation takes the input of a single variable $x_j(s;t)$ and reshapes its distribution using the function $T$ to produce $x^{\text{Trans}}_{j}(s;t)$. When a variable has a skewed distribution, transformations such as log, square root, or cubic root can adjust the distribution towards normality. For example, in the Human Development Index (HDI), a logarithmic transformation is applied to the variable Gross National Income per capita (GNI), to reduce its impact on HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional shape of the variable. It includes techniques such as centering, z-score standardization, and min-max standardization and can be expressed as $[x_{j}(s;t) - \alpha]/\gamma$ where $\alpha$ and $\gamma$ are constants . In the Human Development Index (HDI), the three dimensions (health, education, and economy) are converted into the same scale (0-1) using min-max standardization.

@@

Dimension reduction takes the multivariate information $\mathbf{x}(s;t)$, where $\mathbf{x} \subseteq \mathbb{R}^m \times \mathbb{R}^n$, or a subset of variables in $\mathbf{x}(s;t)$, as the input. It summarises the high-dimensional information into a lower-dimension representation $\mathbf{y}(s;t)$, where $\mathbf{y} \subseteq \mathbb{R}^d$ and $d < p$, as the output. The transformation can be based on domain-specific knowledge, originating from theories describing the underlying physical processes, or guided by statistical methods. For example, the Standardized Precipitation-Evapotranspiration Index (SPEI) [@spei] calculates the difference $D$ between precipitation ($P$) and potential evapotranspiration ($\text{PET}$), using a water balance model ($D = P - \text{PET}$). This is the only step that differs from the Standardized Precipitation Index (SPI), and can be considered to be a dimension reduction using a particular linear combination.

### 2024-04-12 tidyindex.tex
Changes sect 4

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Science, University of Texas at Austin, Austin,
Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{3} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{5} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and the variability therein. Here we
discuss a modular data pipeline recommendation to assemble indexes. It
is universally applicable to index computation and allows investigation
of index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
usage of the pipeline framework: comparison of two different indexes
designed to monitor the spatio-temporal distribution of drought in
Queensland, Australia; the effect of dimension reduction choices on the
Global Gender Gap Index (GGGI) on countries' ranking; and how to
calculate bootstrap confidence intervals for the Standardized
Precipitation Index (SPI). The methods are supported by a new R package,
called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, breakable, interior hidden, sharp corners, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced]}{\end{tcolorbox}}\fi

@@

The spatial processing module takes a single variable with a fixed
temporal dimension, \(x_{tj}(s)\), as input. This step transforms the
variable from the original spatial dimension \(s\) into the new
dimension \(s^\prime \in D_{s^\prime}\) through
\(x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]\). The change of spatial
dimension allows for the alignment of variables collected from different
measurements, such as in-situ stations and satellite imagery, or
originating from different resolutions. This also includes the
aggregation of variables into different levels, such as city, state, and
country scales.

@@

Variable transformation takes the input of a single variable
\(x_j(s;t)\) and reshapes its distribution using the function \(T[.]\)
to produce \(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
distribution, transformations such as log, square root, or cubic root
can adjust the distribution towards normality. For example, in the Human
Development Index (HDI), a logarithmic transformation is applied to the
variable Gross National Income per capita (GNI), to reduce its impact on
HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional
shape of the variable. It includes techniques such as centering, z-score
standardization, and min-max standardization and can be expressed as
\([x_{j}(s;t) - \alpha]/\gamma\). In the Human Development Index (HDI),
the three dimensions (health, education, and economy) are converted into
the same scale (0-1) using min-max standardization.

@@

Dimension reduction takes the multivariate information
\(\mathbf{x}(s;t)\), where \(\mathbf{x} \subseteq \mathbb{R}^p\), or a
subset of variables in \(\mathbf{x}(s;t)\), as the input. It summarises
the high-dimensional information into a lower-dimension representation
\(\mathbf{y}(s;t)\), where \(y \subseteq \mathbb{R}^d\) and \(d < p\),
as the output. The transformation can be based on domain-specific
knowledge, originating from theories describing the underlying physical
processes, or guided by statistical methods. For example, the
Standardized Precipitation-Evapotranspiration Index (SPEI)
(Vicente-Serrano, Beguería, and López-Moreno 2010) calculates the
difference \(D\) between precipitation (\(P\)) and potential
evapotranspiration (\(\text{PET}\)), using a water balance model
(\(D = P - \text{PET}\)). This is the only step that differs from the
Standardized Precipitation Index (SPI), and can be considered to be a
dimension reduction using a particular linear combination.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Science, University of Texas at Austin, Austin,
Texas, USA\\ \textsuperscript{2} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{3} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{4} Guangdong Provincial Key
Laboratory of Interdisciplinary Research and Application for Data
Science, BNU-HKBU United International College, Zhuhai,
Guangdong, China\\ \textsuperscript{5} School of Mathematics and
Statistics, University of Melbourne, Melbourne, Victoria, Australia
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and the variability therein. Here we
discuss a modular data pipeline recommendation to assemble indexes. It
is universally applicable to index computation and allows investigation
of index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
usage of the pipeline framework: comparison of two different indexes
designed to monitor the spatio-temporal distribution of drought in
Queensland, Australia; the effect of dimension reduction choices on the
Global Gender Gap Index (GGGI) on countries' ranking; and how to
calculate bootstrap confidence intervals for the Standardized
Precipitation Index (SPI). The methods are supported by a new R package,
called \texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, enhanced, boxrule=0pt, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, sharp corners]}{\end{tcolorbox}}\fi

@@

The spatial processing module takes a single variable with a fixed
temporal dimension, \(x_{tj}(s)\), as input. This step transforms the
variable from the original spatial dimension \(s\) into the new
dimension \(s^\prime \in D_{s^\prime}\) through
\(x^{\text{Spat}}_{tj}(s^\prime) = g[x_{tj}(s)]\) via a function \(g\).
The change of spatial dimension allows for the alignment of variables
collected from different measurements, such as in-situ stations and
satellite imagery, or originating from different resolutions. This also
includes the aggregation of variables into different levels, such as
city, state, and country scales.

@@

Variable transformation takes the input of a single variable
\(x_j(s;t)\) and reshapes its distribution using the function \(T\) to
produce \(x^{\text{Trans}}_{j}(s;t)\). When a variable has a skewed
distribution, transformations such as log, square root, or cubic root
can adjust the distribution towards normality. For example, in the Human
Development Index (HDI), a logarithmic transformation is applied to the
variable Gross National Income per capita (GNI), to reduce its impact on
HDI, particularly for countries with high GNI values.

@@

Unlike variable transformation, scaling maintains the distributional
shape of the variable. It includes techniques such as centering, z-score
standardization, and min-max standardization and can be expressed as
\([x_{j}(s;t) - \alpha]/\gamma\) where \(\alpha\) and \(\gamma\) are
constants . In the Human Development Index (HDI), the three dimensions
(health, education, and economy) are converted into the same scale (0-1)
using min-max standardization.

@@

Dimension reduction takes the multivariate information
\(\mathbf{x}(s;t)\), where
\(\mathbf{x} \subseteq \mathbb{R}^m \times \mathbb{R}^n\), or a subset
of variables in \(\mathbf{x}(s;t)\), as the input. It summarises the
high-dimensional information into a lower-dimension representation
\(\mathbf{y}(s;t)\), where \(\mathbf{y} \subseteq \mathbb{R}^d\) and
\(d < p\), as the output. The transformation can be based on
domain-specific knowledge, originating from theories describing the
underlying physical processes, or guided by statistical methods. For
example, the Standardized Precipitation-Evapotranspiration Index (SPEI)
(Vicente-Serrano, Beguería, and López-Moreno 2010) calculates the
difference \(D\) between precipitation (\(P\)) and potential
evapotranspiration (\(\text{PET}\)), using a water balance model
(\(D = P - \text{PET}\)). This is the only step that differs from the
Standardized Precipitation Index (SPI), and can be considered to be a
dimension reduction using a particular linear combination.
